Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimization Choices

### 2025-03-04 jso.tex
before Section 6

It is expected that for a PP index with high squint angle, the
optimization (\eqref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
half-point index value improvement percentage provides an appropriate
mathematical definition of squintability, which matches the intuition
behind this concept, while being amenable to numerical computation.

@@

To solve the optimization problem embedded in the PP guided tour, a
starting projection, an index function, the number of jellyfish, and the
maximum number of trials (tries) are provided as input. Then, the
current projection is evaluated by the index function. The projection is
then moved in a direction determined by a random factor, influenced by
how far along we are in the optimization process. Occasionally,
completely new directions may be taken like a jellyfish might with ocean
currents. A new projection is accepted if it is an improvement compared
to the current one, rejected otherwise. This process continues and
iteratively improves the projection, until the pre-specified maximum
number of trials is reached.

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, left=2mm, leftrule=.75mm, opacityback=0, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, toprule=.15mm, breakable, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}]

@@

\begin{quote}
Calculate the time control value, \(c_t\), based on
\texttt{current\_idx} and \texttt{max\_trial}
\end{quote}

@@

\begin{quote}
For each projection, compare with a random jellyfish and adjust towards
or away from it (active)
\end{quote}
\end{quote}

@@

\begin{quote}
\textbf{if} \texttt{trial\_id} \(\ge\) \texttt{max\_trial}, print the
last best projection \textbf{exit}
\end{quote}

@@

This section explains the details of two simulation studies: (1)
comparison between the existing creeping random search (Zhang et al.
2021; Laa and Cook 2020) and JSO, and (2) examining the factors
affecting the performance of JSO.

@@

The CRS is the main optimization routine currently used for the guided
tour. Here the two optimizers are compared on the task of finding the
pipe using the \texttt{holes} index in data with dimensions,
\(d = 6, 8, 10, 12\). Fifty simulated data sets are used for each
dimension. JSO uses 100 jellyfish with a maximum of 100 tries, while the
CRS allows a maximum of 1000 tries at each iteration before the
algorithm terminates. These choices enable fair comparison between CRS
and JSO, that conforms to how they are used in practice.

The performance of the opt misers is measured by the success rate, which
is defined as the proportion of simulations that achieves a final index
value within 0.05 of the best index value found among all 50
simulations. Figure~\ref{fig-success-rate} illustrates why the choice of
0.05 is reasonable: the pipe is not recognizable in the projected data.
This is motivated by Laa and Cook (2020)'s approach to investigating PP
indexes.

The results of the simulation are collected using the data structure
used in Zhang et al. (2021) for assessing PP optimizers. The design
parameters are stored along with index value, projection basis, random
seed, and computation time.

@@

\caption{\label{fig-success-rate}How success rate is calculated,
illustrated using the optimal projections from 50 optimizations of 8D
pipe data sorted by index value. The pipe shape is recognisable in the
projection index values between 0.933-0.969. Of the 50 simulations, 43
achieved an index value within 0.05 of the best, resulting in a success
rate of 0.86.}

@@

To examine the performance of JSO relative to hyper-parameter choices,
the \texttt{holes} index is used to find the 2D pipe in data of
dimensions 4-12. Additional variables are Gaussian noise. The levels of
hyper-parameters are 20, 50, and 100 jellyfish and a maximum of 50 and
100 attempts. Fifty samples of each data set are generated, and the
success rate is calculated in the same manner as described in
Section~\ref{sec-app-1}.

This set of simulations is expanded to assess the performance of JSO
relative to the index properties, with a second data set type and
additional PP indexes. The second data set has a sine wave in two
dimensions and Gaussian noise in the remaining dimensions. Six
additional PP indexes that should be able to extract the sine wave are
used: \texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, following Laa and Cook (2020).
(The supplementary material includes definitions of these PP indexes,
and descriptions of the pipe and sine wave data simulation.). The JSO
hyper-parameters, number of jellies and maximum tries, are also
included. This results in a total of 76 scenarios, comprising of 30
computed on the pipe data and 46 on the sine wave data. The choice of
4-12 dimensions covers the range of easy to relatively difficult for the
optimizers to detect the structured projections. The same approach is
used, that 50 samples are generated and the success rate is calculated
at each level of the simulation design. A success rate of 0 indicates
the optimizer did not come sufficiently close to the optimal projection.

@@

\caption{\label{fig-smoothness}Steps for calculating smoothness in a
projection pursuit problem. Given the target shape, data dimension and
the index function: 1) sample random bases given the orthonormality
contraint and calculate their corresponding index values, and 2) fit a
Gaussian process model of index values against the sampled bases to
obtain the smoothness measure.}

@@

Figure~\ref{fig-smoothness} describes the procedure to compute
smoothness. Index values are calculated for each of 500 simulated random
bases. A Gaussian process model is fitted to the resulting data to
obtain the smoothness measure for the index, as described in
Section~\ref{sec-smoothness}.

@@

\caption{\label{fig-squintability}Steps for calculating squintability in
a projection pursuit problem. Given the target shape, data dimension and
the index function: 1) sample random bases given the orthonormality and
projection distance contraint and calculates their corresponding index
values, 2) interpolate the sampled bases to the optimal basis and
calculate the projection distance and the index value. 3) bin the index
values by projection distances through averaging index value, 4) fit the
scaled sigmoid function in equation \eqref{eq-squintability} to the
binned index values against projection distances using non-linear least
square to obtian the squintability measure using equation
(\ref{eq-squintability-parametric}).}

@@

Figure~\ref{fig-squintability} describes the procedure to compute
squintability. Index values and projection distances are computed for
the projections along interpolated paths between 50 random bases and the
optimal basis. An interpolation step size of 0.005 is used. The
projection distances are binned to 0.005 intervals and index values are
averaged in each bin. The four-parameter scaled logistic function
(\ref{eq-parametric}) is fitted to the index values against projection
distances, estimated by non-linear least squares to obtain the
squintability measure as Equation (\ref{eq-squintability-parametric}).

A generalized linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting success rate.
Predictors are smoothness, squintability, data dimension and JSO
hyper-parameters.

>>>>

It is expected that for a PP index with high squint angle, the
optimization \eqref{eq-optimization} should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
half-point index value improvement percentage provides an appropriate
mathematical definition of squintability, which matches the intuition
behind this concept, while being amenable to numerical computation.

@@

To solve the optimization problem embedded in the PPGT with JSO, a
starting projection, an index function, the number of jellyfish, and the
maximum number of trials (tries) are provided as input. Then, the
current projection is evaluated by the index function. The projection is
then moved in a direction determined by a random factor, influenced by
how far along we are in the optimization process. Occasionally,
completely new directions may be taken like a jellyfish might with ocean
currents. A new projection is accepted if it is an improvement compared
to the current one, rejected otherwise. This process continues and
iteratively improves the projection, until the pre-specified maximum
number of trials is reached.

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, leftrule=.75mm, breakable, coltitle=black, colback=white, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, opacityback=0, left=2mm, bottomtitle=1mm, rightrule=.15mm, toprule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, titlerule=0mm]

@@

\begin{quote}
Calculate the time control value, \(c_t\), based on \texttt{trial\_id}
and \texttt{max\_trial}
\end{quote}

@@

\begin{quote}
For each projection, compare with a random jellyfish in
\texttt{current\_projections} and adjust towards or away from it based
on their corresponding \texttt{current\_idx} (active)
\end{quote}
\end{quote}

@@

\begin{quote}
\textbf{exit}
\end{quote}

@@

This section explains the details of two simulation studies: (1) a
comparison between the JSO and an existing optimizer, creeping random
search (Zhang et al. 2021; Laa and Cook 2020), and (2) a modelling study
to examine how factors (smoothness, squintability, data dimension, and
JSO hyper-parameters) affect its success rate in detecting holes and
sine-wave pattern across a collection of index functions.

@@

The CRS is the main optimization routine currently used for the guided
tour. Here the two optimizers are compared on the task of finding the 2D
pipe structure using the \texttt{holes} index in data at dimensions 6,
8, 10, and 12. Additional variables are Gaussian noise. JSO uses 100
jellyfish with 100 iterations, while CRS allows up to 1000 trials per
iteration terminates if no better projection is found within these
trials. These choices enable fair comparison between CRS and JSO, that
conforms to how they are used in practice. Each task is repeated 50
times to evaluate the performance between the two optimizers.

The performance of the optimizers is measured by the success rate, which
is defined as the proportion of simulations that achieves a final index
value within 0.05 of the best index value found among all 50
simulations. Figure~\ref{fig-success-rate} illustrates why the choice of
0.05 is reasonable: the pipe is not recognizable in the projected data.
This is motivated by Laa and Cook (2020)'s approach to investigating PP
indexes.

The results of the simulation are collected using the data structure
used in Zhang et al. (2021) for assessing PP optimizers. The design
parameters are stored along with index value, projection basis, and
random seed used.

@@

\caption{\label{fig-success-rate}How success rate is calculated,
illustrated using the optimal projections from 50 optimizations of 8D
pipe data sorted by index value. The pipe shape is recognizable in the
projection index values between 0.933-0.969. Of the 50 simulations, 43
achieved an index value within 0.05 of the best, resulting in a success
rate of 0.86.}

@@

To examine the performance of JSO relative to hyper-parameter choices,
the \texttt{holes} index is used to find the 2D pipe in data of
dimensions 4, 6, 8, 10, and 12. The levels of hyper-parameters are 20,
50, and 100 jellyfish and a maximum of 50 and 100 attempts. Each task is
repeated fifty times to calculate the success rate as described in
Section~\ref{sec-app-1}.

This set of simulations is expanded to assess the performance of JSO
relative to the index properties, with a second data set type and
additional PP indexes. The second data set has a sine wave in two
dimensions and Gaussian noise in the remaining dimensions. Seven
additional PP indexes that should be able to extract the sine wave are
used: \texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{splines2d}, \texttt{skinny} and \texttt{stringy2}, following Laa
and Cook (2020). (The supplementary material includes definitions of
these PP indexes, and descriptions of the pipe and sine wave data
simulation.). The JSO hyper-parameters, number of jellies and maximum
tries, are also included. This results in a total of 76 scenarios,
comprising of 30 computed on the pipe data and 46 on the sine wave data.
The choice of 4-12 dimensions covers the range of easy to relatively
difficult for the optimizers to detect the structured projections. The
same approach is used, that each experiment is repeated 50 times and
success rate is calculated at each level of the simulation design.

@@

\caption{\label{fig-smoothness}Steps for calculating smoothness in a
projection pursuit problem: 1) sample random bases given the
orthonormality contraint and calculate their corresponding index values,
2) fit a Gaussian process model of index values against the bases to
obtain the smoothness measure, 3) take the smoothness parameter from the
Gaussian process as the smoothness metric.}

@@

Recall that smoothness and squintability are metrics used to
characterize the index function for PP tasks and are invariant to the
optimizer and its hyper-parameters. Thus simulations with different JSO
hyper-parameters for the same PP task share the same smoothness value.
In total, smoothness and squintability metrics are calculated on the 22
PP tasks in our simulation.

Figure~\ref{fig-smoothness} describes the procedure to compute
smoothness. For each PP task, 500 random bases are simulated and index
values are calculated for each. A Gaussian process model is then fitted
to the resulting data to obtain the smoothness measure for the index, as
described in Section~\ref{sec-smoothness}.

@@

\caption{\label{fig-squintability}Steps for calculating squintability in
a projection pursuit problem: 1) sample random bases given the
orthonormality and projection distance contraint and calculates their
corresponding index values, 2) interpolate the sampled bases to the
optimal basis and calculate the projection distance and the index value.
3) average the index values by projection distances at each 0.005
distance bin, 4) fit the scaled sigmoid function in Equation
\eqref{eq-logistic} and \eqref{eq-parametric} , 5) calculate the
squintability metric using Equation
(\ref{eq-squintability-parametric}).}

@@

Figure~\ref{fig-squintability} describes the procedure to compute
squintability. For each PP task, 50 random bases are generated and
interpolated to the optimal basis with a step size of 0.005. The index
values are averaged over the distance window of 0.005. Index values are
calculated for all the interpolated bases and averaged over the distance
bin of 0.005. The logisitic function in Equation (\ref{eq-logistic}) and
(\ref{eq-parametric}) is fitted with non-linear least squares and
calculate the squintability measure using Equation
(\ref{eq-squintability-parametric}).

A generalized linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting the success rate.
Predictors are smoothness, squintability, data dimension, and JSO
hyper-parameters.
